Two subtle problems with over-representation analysis

### 2024-07-08 manuscript/manuscript.Rmd
bmc bioinfo

---
title: "Two subtle problems with over-representation analysis"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: plos-computational-biology.csl
---

@@

# Abstract

Over-representation analysis (ORA) is used widely to assess the enrichment of functional
categories in a gene list compared to a background list.
ORA is therefore a critical method in the interpretation of 'omics data, relating gene lists
to biological functions and themes.
Although ORA is hugely popular, we and others have noticed two potentially undesired
behaviours of some ORA tools.
The first one we call the "background problem", because it involves the software eliminating
large numbers of genes from the background list if they are not annotated as belonging to any
category.
The second one we call the "false discovery rate problem", because some tools underestimate
the true number of parallel tests conducted.
Here we demonstrate the impact of these issues on several real RNA-seq datasets and use
simulated RNA-seq data to quantify the impact of these problems.
We show that the severity of these problems depends on the gene set library, the number of
genes in the list, and the degree of noise in the dataset.
These problems can be mitigated by changing packages/websites for ORA or by changing to
another approach such as functional class scoring.

# Keywords

@@

# Main Body

## Introduction

@@

To test the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.44.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].

@@

The second problem is potentially more serious in that it could lead to increasing false
positives under specific conditions; when the foreground list is short, and the median pathway
size is relatively small.
These false positives arise due to an underestimate of the actual number of tests conducted.
In extreme cases, correction of this problem can lead to around half of the statistically
significant results changing.

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 1**), but there are also several tools free of these problems.

@@

The purpose of this work was to describe two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, ORA was less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results [@Ziemann2024-ts].
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

@@

# Data and Software Availability

## Underlying data

Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io).

## Software and Code

* The code repository including template Dockerfile and R Markdown script are available on
GitHub (https://github.com/markziemann/background).

* The example Docker image is available on DockerHub
(https://hub.docker.com/r/mziemann/background).

* The code repository and Docker image have been uploaded to Zenodo for long-term archiving
(https://zenodo.org/record/8170984).

## Availability of other materials

@@

No competing interests were disclosed.

## Grant Information

The authors declared that no grants were involved in supporting this work.

@@

# Bibliography

>>>>

---
title: "Two subtle problems with over-representation analysis"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
bibliography: references.bib
csl: bmc-bioinformatics.csl
---

@@

## Abstract

**Background**: Over-representation analysis (ORA) is used widely to assess the enrichment of functional
categories in a gene list compared to a background list.
ORA is therefore a critical method in the interpretation of 'omics data, relating gene lists
to biological functions and themes.
Although ORA is hugely popular, we and others have noticed two potentially undesired
behaviours of some ORA tools.
The first one we call the "background problem", because it involves the software eliminating
large numbers of genes from the background list if they are not annotated as belonging to any
category.
The second one we call the "false discovery rate problem", because some tools underestimate
the true number of parallel tests conducted.

**Results**: Here we demonstrate the impact of these issues on several real RNA-seq datasets and use
simulated RNA-seq data to quantify the impact of these problems.
We show that the severity of these problems depends on the gene set library, the number of
genes in the list, and the degree of noise in the dataset.

**Conclusion**: These problems can be mitigated by changing packages/websites for ORA or by changing to
another approach such as functional class scoring.

## Keywords

@@

## Background

@@

To quantify the effect of these two problems on RNA-seq based enrichment analysis, data from seven
transcriptomic studies were downloaded from DEE2 using the getDEE2 R/Bioconductor package
[@Ziemann2019-ux].
Raw data for these seven studies are available from NCBI Sequence Read Archive under the
accession numbers SRP128998, SRP038101, SRP037718, SRP096177, SRP097759, SRP253951 and
SRP068733 [@Felisbino2021-hc; @Lund2014-ds; @Rafehi2014-rv; @Keating2014-ab;
@Blanco-Melo2020-hh; @Rafehi2017-qn].
For each dataset, kallisto (v0.43.1) [@Bray2016-lu] transcript counts were aggregated to the
gene level.
Genes with fewer than 10 reads per sample on average were removed from downstream analysis.
Remaining genes passing this selection were included in the background gene set.
Differential expression analysis was conducted with DESeq2 v1.44.0 [@Love2014-wg], with
genes identified by their Ensembl identifiers.
Gene symbols were then fetched using biomaRt v2.60.0, based on Ensembl version 112 [@Durinck2009-dr].

@@

The FDR problem is potentially more serious in that it could lead to increasing false
positives under specific conditions; when the foreground list is short, and the median pathway
size is relatively small.
These false positives arise due to an underestimate of the actual number of tests conducted.
In extreme cases, correction of this problem can lead to around half of the statistically
significant results changing.

ClusterProfiler is not the only tool to suffer from these problems, they are widespread
(**Table 1**). Still, there are several tools free of these problems.

@@

The purpose of this work was to describe two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results [@Ziemann2024-ts].
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

@@

## Declarations

### Ethics approval and consent to participate

Not applicable.

### Consent for publication

Not applicable.

### Availability of data and materials

Publicly available data were obtained from Digital Expression Explorer 2 (dee2.io),
using the accession numbers mentioned in the Methods section.
The code repository including template Dockerfile and R Markdown script are available on
GitHub (https://github.com/markziemann/background).
The Docker image is available on DockerHub (https://hub.docker.com/r/mziemann/background).
The code repository and Docker image have been uploaded to Zenodo for long-term archiving
(https://zenodo.org/record/TODOXXX).

The authors declare that they have no competing interests.

## Funding

This work was supported by Burnet Institute and Deakin University.
Authors declare that no grants were involved in supporting this work.

@@

@@

## Bibliography
